---
title: "TrueBaseballworking"
author: "Brenden Bready"
date: "2024-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# Loading necessary libraries
library(baseballr)
library(tidyverse)
```

```{r}
# Reading in our play by play data, the data was scraped in a previous R script
pbp = read.csv("pbpdata2021.csv")
```

# Expected Runs

## 1. Calculating run expectancy

```{r}
# Making cases for different situations to help with looping
pbp = pbp %>%
  mutate(bases = case_when(
    is.na(matchup.postOnFirst.fullName) & is.na(matchup.postOnSecond.fullName) & is.na(matchup.postOnThird.fullName) ~ 0,
    is.na(matchup.postOnSecond.fullName) & is.na(matchup.postOnThird.fullName) ~ 1,
    is.na(matchup.postOnFirst.fullName) & is.na(matchup.postOnThird.fullName) ~ 2,
    is.na(matchup.postOnFirst.fullName) & is.na(matchup.postOnSecond.fullName) ~ 3,
    is.na(matchup.postOnThird.fullName) ~ 12,
    is.na(matchup.postOnSecond.fullName) ~ 13,
    is.na(matchup.postOnFirst.fullName) ~ 23,
    TRUE ~ 123
  ))
```

```{r}
# Filter data to only include first pitch
pitchone = pbp %>%
  filter(pitchNumber == 1)
```


```{r}
# Number of times each situation occured - seems to check out
pbp %>%
  group_by(bases) %>%
  summarise(n())
```

```{r}
# Get scores at end of inning, subtract scores from beginning of inning, divide by number of situations

pbp %>%
  filter(bases == 123 & count.outs.start == 0) %>%
  filter(pitchNumber == 1)

# Getting all of the last at bats of an inning (for final inning scores)
lastbats = pitchone %>%
  group_by(game_pk, about.inning, about.halfInning) %>%
  slice(n())
```

```{r}
# Get scores for current situation
scores = pitchone %>%
  filter(bases == 0 & count.outs.start == 2)

tosub = right_join(lastbats, scores, by = c("game_pk", "game_date", "about.inning", "about.halfInning"))

er = tosub %>%
  ungroup() %>%
  summarise(sum((result.homeScore.x + result.awayScore.x) - (result.homeScore.y + result.awayScore.y))/n())


```

```{r}
# Start by initializing a data frame to store results
expruns = data.frame(outs = integer(), bases = integer(), exruns = double())

# Set the loop to go through each out and base runner combination
for (i in unique(pitchone$count.outs.start)){
  for (j in unique(pitchone$bases)){
    
    # Filter to get current score at each baserunner/out combo
    scores = pitchone %>% 
      filter(bases == j & count.outs.start == i)
    
    # Calculate expected runs by joining the last at bats with each of the scores
    # at the start of the inning, ungroup results, sum and divide by total number of 
    # situations where that event occured
    exprun = right_join(lastbats, 
                       scores,
                       by = c("game_pk", "game_date", 
                              "about.inning", "about.halfInning")) %>%
      ungroup() %>%
      summarise(sum((result.homeScore.x + result.awayScore.x) - 
                      (result.homeScore.y + result.awayScore.y))/n())
    
    # Store results in a data frame and join with others
    result = data.frame(outs = i, bases = j, exruns = exprun)
    expruns = rbind(expruns, result)
  }
}
expruns = expruns %>%
  arrange(outs, bases) 
colnames(expruns)[3] <- "exruns"
```

```{r}
exprunsclean = pivot_wider(expruns, names_from = outs, values_from = exruns)
```

## 2 Runs Created


Can go back and get the data with player names included


filter to first pitch and then count occurances of each batter to get total number of at bats



Number of runs at end of inning - number of runs when they start batting

```{r}
# First, need to find top 10 players in runs created

```

For each play, we can make a column entering the expected runs

Then we can group by the inning and half inning, (new ER - old ER) + (new score - old score)


```{r}
# adding current expected runs as a column to dataset
pitchone = left_join(pitchone, expruns, 
                            by = c("count.outs.start" = "outs", "bases" = "bases"))
```

```{r}
# Runs created during that at bat

pitchone = pitchone %>%
  group_by(game_pk, about.inning, about.halfInning) %>%
  mutate(runscreated = (lead(exruns) - exruns) + 
           (lead(result.homeScore) - result.homeScore) + 
           (lead(result.awayScore) - result.awayScore)) %>%
  ungroup() %>%
  mutate(runscreated = ifelse(is.na(runscreated), 0, runscreated))

totalrunscreated = pitchone %>%
  group_by(matchup.batter.fullName) %>%
  summarise(total_runs_created = round(sum(runscreated), 3), atBats = n()) %>%
  filter(atBats >= 100) %>%
  arrange(desc(total_runs_created)) %>%
  mutate(Player = matchup.batter.fullName) %>%
  select(Player, total_runs_created, atBats)



avgrunscreated = pitchone %>%
  group_by(matchup.batter.fullName) %>%
  summarise(average_runs_created = round(sum(runscreated)/n(), 3), atBats = n()) %>%
  filter(atBats >= 100) %>%
  arrange(desc(average_runs_created)) %>%
  mutate(Player = matchup.batter.fullName) %>%
  select(Player, average_runs_created, atBats)


```

```{r}
# Batters with 100 or more at bats
battersover100 = pitchone %>%
  group_by(matchup.batter.fullName) %>%
  summarise(count = n()) %>%
  filter(count >= 100)

pitchonebatters <- pitchone %>%
  semi_join(battersover100, by = c("matchup.batter.fullName" = "matchup.batter.fullName"))
```

```{r message=FALSE}
fangraphs = read_csv("fangraphs.csv")
colnames(fangraphs)[1] <- "fgrank"
```

```{r fig.width=5, fig.height=4}
# Show difference between my rankings and fangraphs 
avgrunscreated = avgrunscreated %>%
  mutate(myrank = row_number())

ranksjoined = left_join(avgrunscreated, fangraphs, by = c("Player" = "Name"))
plot(ranksjoined$myrank, ranksjoined$fgrank, 
     xlab = "Ranking my average runs created",
     ylab = "Fangraphs WAR ranking",
     main = "My rankings vs fangraphs WAR rankings")
```

```{r fig.width=5, fig.height=4}
# Show difference between my rankings and fangraphs 
totalrunscreated = totalrunscreated %>%
  mutate(myrank = row_number())

ranksjoined2 = left_join(totalrunscreated, fangraphs, by = c("Player" = "Name"))
plot(ranksjoined2$myrank, ranksjoined2$fgrank, 
     xlab = "Ranking my total runs created",
     ylab = "Fangraphs WAR ranking",
     main = "My rankings vs fangraphs WAR rankings")
```





